Error-Centric Annotation of Learner Corpora

### 2009-04-09 text/paper.tex
Paper paper paper
SCISSSOSOOOSOSSORRSRSS!!!

\documentclass[12pt]{scrartcl}
\usepackage[utf8x]{inputenc}
\usepackage{graphicx,linguex,natbib,aleks,amsmath,amsthm}
\author{Magdalena Leshtanska \and Aleksandar Dimitrov}
\title{Error-Centric Annotation of Learner Corpora}
\begin{document}
\maketitle

@@

\tableofcontents
\abstract{I am an abstract!}
\section{Introduction}\label{sec:intro}

Literature overview. Bla Bla.

All of the annotation schemes mentioned above focus on the \textbf{text} itself,
and will often even advise the annotator to \textit{modify} the underlying text
with the annotation. We believe that a more error-centric annotation of the data
can be beneficial.

@@

\begin{figure}
\centering
\includegraphics[scale=.2]{taxonomy}
\caption{Error classification taxonomy.}
\end{figure}

@@

The source files to our corpus consisted of plain text files from the
NOCE-corpus % FIXME cite
of beginning to intermediate Spanish learners of English. Our corpus
format is defined in \ref{def:errorformat}

@@

>>>>

\documentclass[12pt]{article}
\usepackage[utf8x]{inputenc}
\usepackage{graphicx,linguex,natbib,aleks,amsmath,amsthm}
\author{Magdalena Leshtanska \and Aleksandar Dimitrov}
\title{Error-Centric Annotation of Learner Corpora}
\begin{document}
\maketitle

@@

\tableofcontents
\abstract{This paper presents an annotation method for errors in learner
language that strives to be as compact and precise as possible, without
sacrificing on generality or correctness. We develop contextualisation
and underspecification methods for error annotations, based on a flexible annotation-driven markup.}

\section{Introduction}\label{sec:intro}

Recently, there has been a growing interest in learner language corpora and
learner errors analysis (citation). There is a need to
identify errors in text, to make the automatic processing of those texts
feasible\cite{nessel}. Also, the potential of learner errors for educational purposes and language
acquisition insight has become increasingly recognized (citation), hence, many
annotation schemes have been devised to identify and classify learner errors,
such as \cite{negthesis}

We propose a novel approach to annotating learner errors, based on the idea that
an error is an inconsistency with its surroundings, rather .  This basic markup approach can be succesfully
combined with any taxonomy for error ??classification. We have coupled it with a
taxonomy for error classification and used the resulting scheme to annotate a
subset of the NOce corpus (citation), achieving an inter-annotation agreement of
???percent.

@@

\begin{figure}
\centering
\includegraphics[scale=.2]{taxonomy}\label{fig:tax}
\caption{Error classification taxonomy.}
\end{figure}

@@

The source files for our corpus consisted of plain text files from the
NOCE-corpus % FIXME cite
of beginning to intermediate Spanish learners of English. Our corpus
format is defined in \ref{def:errorformat}

@@

\appendix

\section{Annotation Manual}
  At any point, if you advance further down the taxonomy tree through a decision
  you make later on, do not mark the error as the more general category, but
  mark it as the more special category instead.

\subsection{Steps in the annotation process}
\begin{enumerate}
\item Look at each token in turn, assuming all the rest is grammatical.
\begin{itemize}
  \item If the token is not spelled correctly, record a spelling error.
  \item If an erroneous token $\tau_1$ is erroneous because it does not
  harmonize with a token $\tau_2$
  from its context,  mark $\tau_1$ as an error and $\tau_2$ as trigger.
  \item If the utterance is ungrammatical, i.e. it does not constitute a part of
  the language, mark it as a grammar mistake.
  \item If the sentence is utterable under a certain condition, but this
  condition is not met here, (i.e. forbidden by the general context of the
  text) mark it as a contextual error.
  \item Now, choose a subcategory from the branch you have chosen. Refer to the
  graph in \ref{fig:tax}.
  Try to be as specific as you can, otherwise, if not completely sure, choose
  the upper node. Try to see in what way a token can be changed, as to
  fit all the additional context. To find out which branch to choose, follow the
  instructions under \ref{branches}
  \item Repeat the process until all possible errors have been annotated, then
  move to the next token.
\end{itemize}

  \item If the context is too erroneous and this erroneous context part affects
  directly the immediate context of your token, try to find another solution,
  because otherwise you will be building error hypotheses on errors. If it doesn't
  (say, it is a spelling mistake), consider it as though it was correct while
  looking for a directly interacting token as trigger.

\item \textit{Local markup}: Keep it as local as possible, try to mark as
triggers tokens that are grammatically dependent on each other in normal
language production.

\item Go from tokens up to other units, such as phrases, and be minimalistic, unless this makes you lose important information.

\item Sometimes the error and/or the trigger will be a sequence. Take the
minimal unit you can that doesn't lead to loss of information. Always choose
the head of a constituent or compound word when it isn't the whole 
constituent that plays a role in error formation.

\ex.  The lovely, charming and amazing lady \textit{laugh} a lot

Here, only take ``lady'' as a trigger. But in cases like \ref{granad}, your
trigger should be ``to study'', and not only ``to'', because otherwise you'd lose
information.  Maybe the learner doesn't have a problem with prepositions, but with infinitives.

\ex. I go for to study in Granada.\label{granad}

\item Avoid cascading and annotating errors based on corrected errors. Only look at the unaltered text.

\item Annotate every possibility. Since we do not apply cascading rules, one can use weighting algorithms to prefer one error to another in the case of concurrent errors:

\ex. \a.\textit{A} mobile \textbf{phones} can be very useful. 
\b. \textbf{A} mobile \textit{phones} can be very useful.

\item Only correct when necessary, do not be tempted to rephrase a construction,
just because it would sound better or is more common, if there is nothing
explicitly wrong with it.

\item No style errors.

\item No punctuation errors

\item No capitalisation, only if it is something specific to English
(Spelling$\to$Register), such as, say, capitalisation of proper nouns or months.

\item If you have to choose between extremely similar possibilities, choose the
one whose target hypothesis would have the least impact on the text in terms of
phrasal constituents.

\end{enumerate}

\section{Branches in the taxonomy}\label{branches}

\subsection{Collocation vs. Replace/Omission/Redundant}

Verb and preposition collocations (i.e. where a certain verb forces a certain
collocation) are to be marked as Collocations, not as an erroneous word. Words
that occur with plural (\textit{one of}, \textit{between}) are also collocation
bound, and an error in agreement is thus a collocation error.

\subsection{Gerunds}

Gerunds are nominalized noun forms.  Place mistakes related to the formation of
gerunds in Morphology $\to$ Inflection, and errors related to the tense in which a
gerund ought (not) to occur in Grammar $\to$ Verbtense.

See also: Derivational Mistakes

\subsection{Morphology}

Morphological negation and plural formation mistakes often occur in English
learner language. We tag them as Morphology and Morphology $\to$ Inflection instead
of choosing a more specific category, but might extend our scheme in the future.

\subsection{Derivational Mistakes}

If a word is wrongly derived from another, (nominalization, suffixation,
prefixation, etc\ldots) tag it as a morphological derivation mistake.

\subsection{Animate/Inanimate}

Is not covered by the scheme right now. Use Replace $\to$ Pronoun or similar. Might
be added in the future.

\subsection{Subject, Object, Predicate vs Pronoun, noun, verb.}

If a whole constituent is missing or wrong, use the constituent classes. If it's
one word out of the constituent, use the part of speech classes. Again: try to
be as general as possible.

\subsection{Number Agreement vs. Redundancy}

\ex. a dogs

\textit{a} is redundant, but because of number agreement. So you must annotate both,
creating a circular dependency, where an erroneous \textit{dogs} has as error
context \textit{a} and an erroneous \textit{a} has as context \textit{docs}

\subsection{Compound nouns:}

If a compound noun of two or more words is erroneous, annotate only the head noun, carrying the mistake.

\subsection{Word order}
\begin{itemize}
\item \textit{Internal word order.} Words should be shuffled within the marked sequence. 
\ex. Not only I must study $\to$ Not only must I study

\item \textit{External word order.} Specified word needs to be moved somewhere
else, its elements moving together as a unit.
\ex. I the river like $\to$ I like the river

\subsection{Verb Tense}

If the verb tense is wrongly formed, (\textit{he had having a nice time}), it's
a grammatical error. Also, if there is a typical trigger word for verb tenses
(such as \textit{since}, \textit{before}, \textit{if}) and the tense is wrong,
it's a grammatical error. If the tense doesn't fit because of the broader
context, it's a contextual mistake.

\section{Source Code and XML Data}\label{sec:github}

The complete code, as well as an up-to-date version of this paper can be found
online at \texttt{www.github.com/adimit/ll-annotation}. The annotation schema's
DTD is \texttt{AnnotationScheme.dtd} in the root directory of the repository.
